Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2019-12-19 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Include suggestions for outlook (two-step model)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2937 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 3758631ef5650fbaf593d89c4c8d13a9c4698d99

This novel approach to circular regression trees and forests complements the
literature on tree-based circular modeling. \cite{Lund:2002} already
introduced a circular regression tree algorithm where binary splits are
made based on an angular distance measure capturing node
homogeneity. However, this only models changes in the conditional mean
but not the conditional variance or full probabilistic distribution 
%% FIXME: first attempt to include reference Gneiting+Balabdaoui+Raftery:2007
which would allow for also considering uncertainty in forecasts \citep{Gneiting+Balabdaoui+Raftery:2007}.
%%
In contrast, we introduce a distributional tree approach that considers
splits in all distribution parameters and yields a full probabilistic predictive model.
In addition, as a natural extension, ensembles or forests of circular
regression trees are presented in this study, showing that these can further
improve predictive power.

@@

The performance of the novel circular trees and forests is assessed in a 
case study for shortterm probabilistic wind direction forecasting at two
airports with different site characteristics. As benchmark models,
probabilistic climatology and persistency models, as well as a state-of-the-art
circular GLM-type model are evaluated based on proper scoring rules. In summary, the
circular trees and forests have the highest predictive performance in this
setting. For cases without changes in the wind regime, lagged response values
provide already highly skillful estimates leading to a good performance of the
persistency model as observed for shortterm wind direction forecasts in this
study. While in these cases the trees and forests also benefit from the highly
informative lagged response, they can compensate for a lower information of
this covariate by incorporating other quantities and possible interactions of
these, contrary to the persistency model (see
Figure~\ref{fig:boxplot_crpsskill}). Hence, the tree-based models provide
reliable forecasts in all tested meteorological settings. For operational use,
a possible extension could be the incorporation of numerical weather
predictions as (additional) covariates. While this probably only slightly
improves the predictive skill for short leadtimes, it possibly extends the
potential forecast range of the different methods. 
%% FIXME: first draft to include reference on wind speed (Thorarinsdottir+Gneiting:2010)
For the specific case of wind direction forecasting it has to be considered
that this quantity is highly related to wind speed. For example, low wind speeds
usually leads to a higher fluctuation in wind direction. One way how to account for
this influence is by modeling both quantities in a bivariate model. %% FIXME: add reference?
Alternatively, the approaches of \cite{Thorarinsdottir+Gneiting:2010} could be 
incorporated in circular regression trees and forests.
%%
Another possible
improvement for obtaining more parsimonious circular trees is to consider
splitting circular covariates into two circle segments by searching two split
points simultaneously rather than sequentially at different depths. While this
might slightly improve the predictive performance of circular trees, this
should not affect the performance of circular forests, as these consist of
very large trees with many different splits.

>>>>

This novel approach to circular regression trees and forests complements the
literature on tree-based circular modeling. \cite{Lund:2002} already
introduced a circular regression tree algorithm where binary splits are
made based on an angular distance measure capturing node
homogeneity. However, this only models changes in the conditional mean
but not the conditional variance or full probabilistic distribution 
%% FIXME: first attempt to include reference Gneiting+Balabdaoui+Raftery:2007/ changed to Gneiting:2008
which would allow for also considering uncertainty in forecasts \citep{Gneiting:2008}.
%%
In contrast, we introduce a distributional tree approach that considers
splits in all distribution parameters and yields a full probabilistic predictive model.
In addition, as a natural extension, ensembles or forests of circular
regression trees are presented in this study, showing that these can further
improve predictive power.

@@

The performance of the novel circular trees and forests is assessed in a 
case study for shortterm probabilistic wind direction forecasting at two
airports with different site characteristics. As benchmark models,
probabilistic climatology and persistency models, as well as a state-of-the-art
circular GLM-type model are evaluated based on proper scoring rules. In summary, the
circular trees and forests have the highest predictive performance in this
setting. For cases without changes in the wind regime, lagged response values
provide already highly skillful estimates leading to a good performance of the
persistency model as observed for shortterm wind direction forecasts in this
study. While in these cases the trees and forests also benefit from the highly
informative lagged response, they can compensate for a lower information of
this covariate by incorporating other quantities and possible interactions of
these, contrary to the persistency model (see
Figure~\ref{fig:boxplot_crpsskill}). Hence, the tree-based models provide
reliable forecasts in all tested meteorological settings. For operational use,
a possible extension could be the incorporation of numerical weather
predictions as (additional) covariates. While this probably only slightly
improves the predictive skill for short leadtimes, it possibly extends the
potential forecast range of the different methods. 
%% FIXME: first draft to include reference on wind speed (Thorarinsdottir+Gneiting:2010)
For the specific case of wind direction forecasting it has to be considered
that this quantity is highly related to wind speed. For example, low wind
speeds usually lead to a high fluctuation in wind direction. Hence, to account
for both quantities simultaneously, distributional tree-based models could be
built for wind vectors employing the bivariate normal distribution as performed
for additive regression models by \citet{Lang+Mayr+Stauffer:2019}.
Alternatively, also accounting for wind speed and direction but extending the
presented methodology, circular trees and forests could be combined with
distributional models for wind speed employing, e.g., the truncated normal
distribution \citep{Thorarinsdottir+Gneiting:2010}. This could be pursued
either in a two-step approach with two independent tree-based distributional
models or in a more sophisticated approach employing a mixture response
distribution.
%One way how to account for
%this influence is by modeling both quantities in a bivariate model. %% FIXME: add reference?
%Alternatively, the approaches of \cite{Thorarinsdottir+Gneiting:2010} could be 
%incorporated in circular regression trees and forests.
%%
Another possible
improvement for obtaining more parsimonious circular trees is to consider
splitting circular covariates into two circle segments by searching two split
points simultaneously rather than sequentially at different depths. While this
might slightly improve the predictive performance of circular trees, this
should not affect the performance of circular forests, as these consist of
very large trees with many different splits.
